Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis

### 2018-04-11 jhir/latex-manuscript/manuscript.tex
update manuscript

\date{Received: 12/03/2018 Revised: 04/15/2018 Accepted: }
% The correct dates will be entered by the editor

@@

\begin{abstract}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking “MI-consistent” counselor behaviors and patient motivational statements (i.e., “change talk”), the specific counselor communication behaviors effective for eliciting patient change talk are a subject of ongoing research. A significant barrier to this research is the time and resource-intensive cognitive tasks, including the manual coding and sequential analysis of coded transcripts, that are an integral part of the qualitative analysis of patient-counselor communication. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks. In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, and closed frequent pattern mining, an algorithm used to mine the closed frequent patterns in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had previously been annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify meaningful and interpretable counselor communication strategies to guide clinical practice.

\keywords{sequential analysis \and hidden markov model \and closed frequent pattern mining \and motivational interviewing \and weight loss \and adolescents}
\end{abstract}

\section{Introduction}
\label{sec:intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change \cite{miller2013motivational}. The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills \cite{miller2013motivational}. The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the “spirit of MI”, an optimal therapeutic state to explore behavior change. MI’s technical hypothesis \cite{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”. Change talk are patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Previous studies \cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention \cite{walker2011influence}. In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk \cite{magill2014technical}. These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.

@@

Several prior works have reported the results of adopting machine learning methods, such as topic models [18-22], classification methods [23-26] and neural networks [23, 27, 28] to the tasks of annotating MI transcripts and primarily for assessing intervention fidelity. Perez-Rosas et al. [25] developed a system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews. This system employed a support vector machine (SVM) classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features, which achieved nearly 90\% accuracy for predicting counselor questions and reflections. Patient’s engagement, verbal and non-verbal accommodation as well as topics discussed during the MI session, were analyzed to identify linguistic and acoustic features of counselor empathy. These features were then used to build a counselor empathy classifier that achieved an accuracy of up to 80\%. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes [23]. The SVM model with the proposed features achieved 75\% accuracy for automated annotation of 17 behavior codes, which is comparable to human coders. In a follow up study [29], we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk). RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews. This paper represents the next phase in this line of research. We examine the efficacy of Hidden Markov Models (HMMs) and Frequent Pattern Mining in identifying the counselor communication strategies antecedent to patient change talk.

HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a “hidden” state, where the hidden state corresponds to a different distribution over observations (i.e., distribution over behavior codes). Sequences of observations are modeled as transitions between different hidden states and sampling observations from each hidden state. HMMs were originally proposed for speech recognition [30], in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling [31, 32]. For example, Doppler ultrasound was used to extract features from imaging data and then an HMM-based classifier was applied to distinguish healthy patients from those with heart disease [31], while HMM was used to capture important characteristics of protein families [32]. In the application of HMM to patient-counselor communication, hidden states correspond to sets of related behavior codes, such as a patient’s underlying motivational state, during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change [33]) as the antecedents of patient’s behavior change [3], there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that likely to lead to patient change talk and commitment language.

@@

In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI’s technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper provides the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al. [47] used HMM and adjusted regression models to examine the association of different sequences of “change talk” utterances with drinking outcomes in a brief motivational intervention. Their HMM model limited hidden states to three, which are "towards change", "away from change" and "non-determined". First and last hidden states decoded for a communication sequences, were used in conjunction with those three hidden states variables as predictors in a bivariate regression model, the results of which indicated that only the last state was significantly associated with drinking 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study to evaluate the utility of using HMM and Frequent Pattern Mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order Markov chain based methods most typically used in MI research. First-order Markov chain models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communication during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.

\section{Methods}
\label{sec:methods}
\subsection{Dataset}
\label{subsec:dataset}
The data set for this study consists of 37 transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss treatment. Each transcript represented a one-on-one interaction between the adolescent patient and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE) [13]. The MYSCOPE is an adaptation of the original MI-SCOPE [48], a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI) [49], the MI Skill Code (MISC) [50], and Amrhein’s conceptualization of change talk and commitment [51]. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% (n = 7) of the sessions for inter-rater reliability which was good (k = .696). The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\subsection{Data preprocessing}
\label{subsec:datapreprocessing}
Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences. For each MI transcript, the stream of behavior codes from beginning to end was analyzed. Successful sequences were those that resulted in a patient change talk or commitment language statement and all preceding behavior codes were included. Unsuccessful sequences were similarly created for sequences resulting in sustain talk. A total of 1,360 sequences were generated. The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which the initial motivation for behavior change is typically high. Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.

\subsection{Data Analysis}
\label{subsec:dataanalysis}
The Hidden Markov Model (HMM; implemented in the hmmlearn package which is publicly available at http://hmmlearn.readthedocs.io/en/latest/index.html) analysis aimed to identify clusters of related behavior codes within successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters. Given a set of behavior code sequences, the posterior inference of HMM parameters involves the deduction of a temporal sequence of hidden states that best explains observations in each sequence. The emission probability matrix describes the distribution of observation symbols (i.e., the MYSCOPE behaviors displayed) for each hidden state and the transition probability matrix describes the transitions among the hidden states.

Training an HMM with a given number of hidden states (N) involves estimating the following parameters using the Baum-Welch algorithm:
M is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table 1).
T is an  state transition probability matrix where  
E is an  matrix where , with observation symbol probability distribution for each state (emission probability matrix), and
p is the initial state distribution vector where .

In this study, two HMM models were trained, one using successful sequences and the other using unsuccessful sequences, with the objective of maximizing log-likelihood on the observed sequence. The optimal number of hidden states was determined by estimating the Bayesian information criterion (BIC) of HMM models with a different number of hidden states and selecting the model with the smallest BIC value, which indicates an optimal fit of HMM model. BIC minimizes negative log-likelihood but penalizes complexity (i.e. \# of hidden states), which is important for avoiding overfitting problem. HMM analyses suggested 5 hidden states was optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure~\ref{fig:bic}).

@@

A second experiment was conducted with Frequent Pattern Mining. Orange3 and arules are two available pattern mining libraries implemented in python and r-programming, which support only a few pattern mining algorithms including frequent itemsets and association rules mining. However, we conducted our experiment using SPMF [52, 53], which offered more than 150 data mining algorithms. Moreover, SPMF is an open-source, fast and lightweight, a most popular and widely used Java-based data mining library. A frequent itemset is determined by surpassing a user-specified threshold called minimum support count\footnote{Support count: \# of times an itemset appears in the database}. For example, \{A\}, \{C\}, {D}, and {C, D} are frequent itemset in the database illustrated in Figure~\ref{fig:fpm}, because these itemsets appear at least 2 times which was the minimum support count identified in the example database. We reported closed frequent itemsets among all sequences of behavior codes in successful and unsuccessful sequences. A frequent itemset is closed if none of its supersets have the same support count [45], where a set X is a superset of another set Y if X contains all elements of the set Y. For example, the itemsets {A}, {C}, {D}, and {C, D} are not closed frequent itemsets because their supersets \{A, B\}, {B, C}, {B, D}, and {B, C, D} have the same number of support. Therefore, {B}, {A, B}, {A, D}, {B, C}, {B, D}, and {B, C, D} are closed frequent itemsets because none of their supersets have same the number of support count. On the other hand, {A, C}, {A, B, C}, {A, B, D}, {A, C, D} and {A, B, C, D} have support count 1, 0, 1, 0, and 1, respectively, which are less than minimum support count and identified as non-frequent itemsets. This study applied FPClose [46], an efficient close frequent pattern mining algorithm in SPMF, to identify frequent patient-counselor behaviors in MI communication sequences. To improve the model’s ability to identify frequent patient behaviors, FPClose [46], a FP-tree based algorithm [54, 55] was used. FPClose identifies closed frequent itemsets by utilizing an FP-tree data structure in combination with the FP-array technique to avoid the frequent traversal of FP-trees. Before applying the FPClose method, the dataset was transformed into the expected format used by the algorithm in the SPMF library. Minimum support threshold is a certain proportion having a meaningful representation of patterns determined by the domain expert. Since there is no standard for the selection of minimum support threshold, by following the studies [56-57], the minimum support count considered was 10\% of the total number of sequences which is 110 for successful and 25 for unsuccessful sequences. In our study, support counts representing ≥10\% of the patterns was the threshold for the result interpretation.

\begin{figure*}
\includegraphics[width=0.95\textwidth]{figures/fpm.eps}
\caption{Different type of itemsets is illustrated with a frequent pattern mining method applied on transaction database by considering minimum support count 2, where TID represents the transaction ID of the corresponding items.}
\label{fig:fpm}
\end{figure*}

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model, are reported in Tables~\ref{tab:transition} and~\ref{tab:emission}. The three behaviors with the highest frequency, representing 45-69\% of each state’s emission transitions, were used to interpret the emission matrix and label the hidden states. Most (84\%) successful sequences began in a state characterized as “high motivation” as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%). Successful sequences began in a state of “high receptivity” 11\% of the time. “High receptivity” sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%). Few successful sequences began in states of “moderate receptivity” and “low receptivity” (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. Moderate receptivity sequences were distinguished from low receptivity sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient’s autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the “active feedback” state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%). Successful sequences transitioned from “high motivation” to “active feedback” most often (41\%). “active feedback”, in turn, most frequently transitioned to “high receptivity” (39\%). The “moderate receptivity” state transitioned to most often to “active feedback” (27\%) and automatically transitioned back to “moderate receptivity” (24\%) or to “low receptivity” (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

In contrast, the majority of unsuccessful sequences (98\%) began in a state of “ambivalence” as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%). About 2\% of the time unsuccessful sequences started in a state of “avoidance”. Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements, and counselor patient-centered information (18\%) distinguished “avoidant” sequences. Both “ambivalent” (84\%) and “avoidant” (61\%) states most frequently transitioned to the “avoidant” state.

@@

Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences are shown in Table~\ref{tab:patterns}. Results from frequent pattern mining analysis indicate that reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%). For each pattern, the significance of difference between successful and unsuccessful communications was reported by computing p-value using Pearson chi-square test. Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p<0.0001), statements emphasizing the patient’s decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423). In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= <0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p=<0.0001), providing information (28.7\% versus 22.1\%, p= 0.0254) and other reflections (11.6\% versus 0\%, p=<0.0001). In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in unsuccessful sequences. In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in successful sequences.

@@

\bibliographystyle{natbib}       % JHIR format
\bibliography{references}   % name your BibTeX data base

\end{document}

>>>>

\date{Received: 3 December 2018 /Revised: 15 April 2018 /Accepted: }
% The correct dates will be entered by the editor

@@

\begin{abstract}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research. A significant barrier to this research is the time and resource-intensive cognitive tasks, including the manual coding and sequential analysis of coded transcripts, that are an integral part of the qualitative analysis of patient-counselor communication. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks. In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients’ motivational statements to guide clinical practice.

\keywords{hidden markov model \and closed frequent pattern mining \and motivational interviewing \and pediatric obesity}
\end{abstract}

\section{Introduction}
\label{sec:intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change [1]. The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills [1]. The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change. MI’s technical hypothesis [2] states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''. Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Previous studies [3] have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention [4]. In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk [5]. These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.

@@

Several prior works have reported the results of adopting machine learning methods, such as topic models [18-22], classification methods [23-26] and neural networks [23, 27, 28] to the tasks of annotating MI transcripts and assessing intervention fidelity. Perez-Rosas et al. [25] developed a system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews. This system employed a support vector machine (SVM) classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features, to predict counselor questions and reflections, which achieved nearly 90\% accuracy. The fidelity is assessed by computing a ratio of questions to reflections, since in MI sessions the goal is to have this ration at least 1:1. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes [23]. The SVM model with the proposed features achieved 75\% accuracy for automated annotation of 17 behavior codes, which is comparable to human coders. In a follow up study [29], we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk). RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews. This paper represents the next phase in this line of research. We examine the efficacy of Hidden Markov Models (HMMs) and Frequent Pattern Mining in identifying the counselor communication strategies antecedent to patient change talk.

HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition [30], in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling [31, 32]. For example, Doppler ultrasound was used to extract features from imaging data and then an HMM-based classifier was applied to distinguish healthy patients from those with heart disease [31], while HMM was used to capture important characteristics of protein families [32]. In the application of HMM to patient-counselor communication, hidden states correspond to sets of related behavior codes, such as a patient’s underlying motivational state, during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change [33]) as the antecedents of patient’s behavior change [3], there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that likely to lead to patient change talk and commitment language.

@@

In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI’s technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper provides the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al. [47] used HMM and adjusted regression models to examine the association of different sequences of ``change talk'' utterances with drinking outcomes in a brief motivational intervention. Their HMM model limited hidden states to three, which are "towards change", "away from change" and "non-determined". First and last hidden states decoded for a communication sequences, were used in conjunction with those three hidden states variables as predictors in a bivariate regression model, the results of which indicated that only the last state was significantly associated with drinking 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study to evaluate the utility of using HMM and Frequent Pattern Mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order Markov chain based methods most typically used in MI research. First-order Markov chain models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.

\section{Materials and Methods}
\label{sec:materialsAndMethods}
\subsection{Dataset}
\label{subsec:dataset}
The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE) [13]. The MYSCOPE is an adaptation of the original MI-SCOPE [48], a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI) [49], the MI Skill Code (MISC) [50], and Amrhein’s conceptualization of change talk and commitment [51]. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\subsection{Data preprocessing}
\label{subsec:datapreprocessing}
Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences. For each MI transcript, the stream of behavior codes from beginning to end was analyzed. Successful sequences are defined as those that resulted in a patient change talk or commitment language statement. Unsuccessful sequences were similarly created for sequences resulting in sustain talk. A total of 1,360 sequences were generated using this approach. The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which the initial motivation for behavior change is typically high. Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.

\subsection{Data Modeling}
\label{subsec:dataModeling}
\subsubsection{Hidden Markov Model}
\label{subsubsec:hiddenMarkovModel}
We applied Hidden Markov Model (HMM)\footnote{we used the implementation in the hmmlearn package publicly available at http://hmmlearn.readthedocs.io/} to identify clusters of behavior codes corresponding to successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters. Given a set of behavior code sequences, the posterior inference of HMM parameters involves the deduction of a temporal sequence of hidden states that best explains observations in each sequence. The rows in the emission probability matrix correspond to the distribution of observation symbols (i.e., the MYSCOPE behaviors displayed) for each hidden state and the transition probability matrix describes the transitions between the hidden states.

Training an HMM with a given number of hidden states ($N$) involves estimating the following parameters using the Baum-Welch algorithm:

\begin{itemize}
\item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
\item $T$ is an state transition probability matrix, in which is the probability of HMM transitioning from state  to state  
\item $E$ is an emission probability matrix, in which is the probability of observing symbol, when HMM is in state 
\item $p$ is the initial state distribution vector where  is the probability of the first HMM state
\end{itemize}

We trained two HMM models, one using all successful sequences and the other one using all unsuccessful sequences. Each model was trained with the objective of maximizing the log-likelihood of all observations in the corresponding set of sequences. The optimal number of hidden states was determined by estimating the Bayesian information criterion (BIC) of HMM models with a different number of hidden states and selecting the model with the smallest value of BIC, which takes into account both log-likelihood and a penalty term for the number of parameters in the model to avoid overfitting. Experiments with different number of hidden states in HMMs estimated on successful and unsuccessful sequences indicated that 5 hidden states were optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure~\ref{fig:bic}).

@@

\subsubsection{Frequent Pattern Mining}
\label{subsubsec:frequentPatternMining}
We applied Frequent Pattern Mining to identify frequently occurring patterns of temporally ordered patient-provider behavior codes in successful and unsuccessful communication sequences. Behavior codes in these patterns may be separated by one or more other codes. For this purpose, we utilized FPClose [46], an efficient state-of-the-art close frequent pattern mining algorithm implemented in SPMF [52, 53], to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences. SPMF is an open-source library providing more than 150 data mining algorithms. Popular non-close frequent pattern mining algorithms include Apriori [35] and FP-Growth [54]. A \textit{frequent pattern} is defined as a pattern of temporally ordered observations, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}. For example, \{A\}, \{C\}, \{D\}, and \{C, D\} are frequent patterns in the example set of sequences in Figure 2, since these patterns appear at least 2 times, which is the minimum support count in this example. In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences. A \textit{frequent pattern is closed} if none of its supersets have the same support count [45], where a set X is a superset of another set Y, if X contains all the elements of the set Y. For example, the itemsets \{A\}, \{C\}, \{D\}, and \{C, D\} in Figure 2 are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\}, and \{B, C, D\} have the same support count. Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\}, and \{B, C, D\} are closed frequent patterns, since none of their supersets have the same support count. On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0, and 1, respectively, which is less than the minimum support count, and thus are identified as non-frequent itemsets. Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we follow the procedure in [55-56] and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.

\begin{figure*}
\includegraphics[width=1.0\textwidth]{figures/fpm.eps}
\caption{Different type of itemsets is illustrated with a frequent pattern mining method applied on transaction database by considering minimum support count 2, where TID represents the transaction ID of the corresponding items.}
\label{fig:fpm}
\end{figure*}

@@

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. The three behaviors with the highest frequency, representing 45-69\% of each state’s emission probability mass, were used to interpret the emission matrix and label the hidden states. Most (84\%) successful sequences began with a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%). Successful sequences began with a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{Moderate receptivity} sequences were distinguished from \textit{low receptivity} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient’s autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} state transitioned to most often to \textit{``active feedback''} (27\%) and automatically transitioned back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

In contrast, the majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%). About 2\% of the time unsuccessful sequences started in a state of \textit{``avoidance''}. Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements, and counselor patient-centered information (18\%) distinguished \textit{``avoidant''} sequences. Both \textit{``ambivalent''} (84\%) and \textit{``avoidant''} (61\%) states most frequently transitioned to the \textit{``avoidant''} state.

Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences are shown in Table~\ref{tab:patterns}. Results from frequent pattern mining analysis indicate that reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%). For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson’s chi-square test. Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p<0.0001), statements emphasizing the patient’s decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423). In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= <0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p=<0.0001), providing information (28.7\% versus 22.1\%, p= 0.0254) and other reflections (11.6\% versus 0\%, p=<0.0001). In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in unsuccessful sequences. In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in successful sequences.

@@

\bibliographystyle{natbib}       % JHIR format
\bibliography{references}   
\end{document}
